Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-12-28 supplement.tex
edits to supplement

Fixing these parameters (i.e., East Africa-Eurasian split at 50ka, Neanderthal
admixture at 45ka) resulted in an estimated value for the split between the
ancestors of East and West African populations at around $60$ka. This timing
agrees with previous LD-based estimates of West African-Eurasian divergence
times \citep{Ragsdale2019-nt}, which used different populations than those used
in this study. A date of $60$ka is also in close agreement with inferred
divergence times from an SFS-based demographic inference (of $\sim65$ka), using
synonymous variation from a subset of populations used in this study (Supp.
Figure~\ref{fig:supp-OOA-SFS}) \aprcomment{TODO: add this figure}.
We therefore took this value ($60$ka) as fixed
in most subsequent analyses. We discuss the sensitivity of our model to these
assumptions in Section~\ref{sec:change-dates}.

@@

All supplementary inference results are provided in Demes format
\citep{Gower2022-yn} as a supplemental zipped archive. For each scenario, we
refit the continuous migration and multiple merger models, with stem migration.

@@

When we allowed the out-of-Africa expansion and Eastern/Western African
divergence times to be jointly fit in model optimization, we found that the
best-fit inferred E/W divergence time and the OOA expansion were more recent
(35--40ka, 45--55ka, resp.). The remainder of the parameters were largely
unaffected by these shifts to more recent divergence times, and our
interpretation of the inferred early history remains unchanged. The $95\%$
confidence intervals around these parameters were relatively narrow ($\approx
5$ka for the OOA split, $\approx 10$ka for the East/West African split),
although this uncertainty does not capture uncertainty due to model
mis-specification (as discussed in Section~\ref{sec:models}).

@@

\aprcomment{Here, addressing major comment 3 from reviewer 4, migration between
Nama and Mene and Nama and GBR.}

@@

\aprcomment{Minor comment 3 from reviewer 4.}

@@

>>>>

Fixing these parameters (i.e., East Africa-Eurasian split at 50ka, Neanderthal
admixture at 45ka) resulted in an estimated value for the split between the
ancestors of East and West African populations at around $60$ka. This timing
agrees with previous LD-based estimates of West African-Eurasian divergence
times, of $\sim61$ka ($50-64$ka, $95\%$ confidence interval)
\citep{Ragsdale2019-nt}, which used a different set populations than those in
this study. A date of $60$ka is also in close agreement with inferred
divergence times of $\sim65$ka from both SFS-based demographic inference, using
synonymous variation from a subset of populations used in this study (Mende,
British, and Han Chinese), as well as that of LD-based inference using these
same groups, of $53$ka ($43-63$ka, $95\%$ confidence interval). Taken
altogether, we decided to fix this value ($60$ka) in most subsequent analyses.
We discuss the sensitivity of our model to these assumptions in
Section~\ref{sec:change-dates}.

@@

All supplementary inference results discussed in this section are provided in
Demes format \citep{Gower2022-yn} as a supplemental zipped archive. For each
scenario, we refit the continuous migration and multiple merger models, with
stem migration.

@@

When we allowed the out-of-Africa expansion and Eastern/Western African
divergence times to be jointly fit in model optimization, we found that the
best-fit inferred E/W divergence time and the OOA expansion were more recent
($45-55$ka and $35-40$ka, resp.). The remainder of the parameters were largely
unaffected by these shifts to more recent divergence times, and our
interpretation of the inferred early history remains unchanged. The $95\%$
confidence intervals around these parameters were relatively narrow ($\approx
5$ka for the OOA split, $\approx 10$ka for the East/West African split),
although this uncertainty does not capture uncertainty due to model
mis-specification (as discussed in Section~\ref{sec:models}).

@@

In preliminary analyses, we found that inferred migration rates between the
Nama and non-East African groups were orders of magnitude smaller than
migration rates between other populations included in our inferences.
Therefore, to reduce the number of parameters needing to be fit we fixed those
rates to zero in subsequent fits, and our reported best-fit models did not
include migration between the Mende and Eurasians, nor between the Nama and
Eurasian populations (aside from recent colonial-period admixture). To assess
the impact of this choice, we re-inferred four of our reported models (single
origin, continuous migration, and merger with and without stem migration),
while allowing for additional parameters to include continuous, symmetric
migration for the duration of the coexistence of the Nama and Mende of the Nama
and British.

Indeed, the inferred migration rates for both of these pairs of populations
were small, with $2N_em \ll 1$ in all cases. In the single origin model, the
Mende-British migration rate was 1.04E-6 ($2N_em = 0.02$) and the Nama-British
rate was 2.67E-6 ($2N_em = 0.05$). In the continuous migration model, these
rates were 2.69E-6 ($2N_em = 0.038$) and 3.38E-6 ($2N_em = 0.048$), resp. In
the merger without stem migration model, these rates were 2.96E-6 ($2N_em =
0.064$) and 5.46E-6 ($2N_em=0.119$), resp. And in the merger with stem
migration model, these rates were 2.64E-6 ($2N_em = 0.055$) and 3.19E-6 ($2N_em
= 0.066$), resp. In comparison with other migration rates, with $2N_em \gtrsim
1$, these migrations have negligible impact on underlying statistics. The fits
to other parameters were negligibly different from those reported with these
migrations fixed to zero, and log-likelihoods were unimproved.

@@

We used observed $\pi_2$ and pairwise diversity in the Mende to normalize two-
and one-locus statistics (see Section~\ref{sec:supp-ld-stats}). To confirm that
this choice of normalization did not impact inferred model parameters, we reran
each of our reported models using data normalized by $\pi_2$ and pairwise
diversity in the Gumuz. As expected, the best fit parameters from fits using
the Gumuz-normalized data were very close to those reported in
Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration},
and they were all well within the reported confidence intervals. Our results
are therefore unlikely to be sensitive to this choice of normalization.

Reinferred demographic models from this and preceding sections are reported in
Demes format \citep{Gower2022-yn} as a supplemental archive.

@@

%\begin{table}[ht]
%    \caption{
%        \label{tab:supp-OOA-SFS}
%        \textbf{OOA parameters using LD and SFS statistics.}
%        \aprcomment{To justify 60ka}
%    }
%    \centering
%    \begin{tabular}[t]{rcc}
%        \toprule
%        Parameter & SFS from synonymous variation & LD from intergenic variation \\
%        \midrule
%        $N_e$ &  & \\
%        $N_{AF}$ &  & \\
%        $N_{B}$ &  & \\
%        $N_{EU0}$ &  & \\
%        $N_{EUF}$ &   & \\
%        $N_{AS0}$ &  & \\
%        $N_{ASF}$ &  & \\
%        $m_{AF-B}$  & & \\
%        $m_{AF-EU}$  & & \\
%        $m_{AF-AS}$  & & \\
%        $m_{EU-AS}$  & & \\
%        $T_{EXP}$ &  & \\
%        $T_{OOA}$ &  & \\
%        $T_{EU-AS}$  & & \\
%        \bottomrule
%    \end{tabular}
%\end{table}
